### 2022-05-23 src/ms.tex
minor updates to introduction

\section{Gravitational microlensing}
\emph{``Do not bodies act upon light at a distance and by their action bend its rays,
    and is not this action strongest at the least distance?''} asked \citet{Newton1704}.
Much later in 1912, even before he published his theory of General Relativity, Einstein
predicted the that a distant light source could be magnified by a foreground mass
relative to an observer on Earth if the two objects were very precisely aligned
\citep{Renn1997}.
At the time Einstein concluded that this \emph{gravitational lensing} effect was
unobservable and did not even bother
publishing the result until 24 years later a young amateur scientist R.W. Mandl convinced
him to write a short paper on the subject \citep{1936Sci....84..506E}.
In that paper he
stated that \emph{``Of course, there is no hope of observing this phenomenon directly.
    First, we shall scarcely ever approach closely enough to such a central line.
    Second, [the angles] will defy the resolving power of our instruments''}.
Still, the consensus at the time was that this was unobservable.

The first definitive observation came in 1979 by \citet{1980ApJ...241..507Y}
who observed a double image of the quasar Q0957+561 and concluded that the two
images correspond to the same object whose light was distorted by a foreground
gravitational lens in the form of a massive galaxy. That the same effect could
be used to detect the presence of planets orbiting around their host stars was
first theorized by \citet{1964PhRv..133..835L} who wrote that \emph{``the
    primary effect of planetary deflectors bound to stars other than the Sun will
    be to slightly perturb the lens action of these stars''} although he was also
sceptical about the possibility of detection, saying that \emph{``associated
    pulses would be so weak and infrequent and of such fleeting duration – perhaps
    a few hours – as to defy detection''}. Gravitational lensing as a method for
discovering exoplanets really took off with the work of Paczyński
\citep{1986ApJ...304....1P,1986ApJ...301..503P,1991ApJ...374L..37M} who also
coined the term \emph{microlensing}, which refers to gravitational lensing in a
regime where the images of the background object cannot be resolved but one can
nevertheless measure its magnification as a function of time.

@@

Microlensing as a method for detecting exoplanets has some unique aspects.
First, it's a one-off event which happens on a timescale of a few minutes up to
several months depending on the distances to the background star and the
lensing star and the mass of the lens. In addition to the fact that there's
only one chance to observe such an event for it to happen at all requires
extremely precise alignment between the lens and the background star and the
chance of observing a stellar microlensing event is about one-in-a-million
(CITE) for a typical star within the Milky Way. To observe a planetary signal
is about an order of magnitude less likely than that. Hence, obtaining a decent
sample of planetary events requires continuous monitoring on the order of
$10^{8}$ stars at the time which way microlensing observations that to focus on
the densest region of the Milky Way -- the galactic bulge.
Figure~\ref{fig:crowded_field} a picture from of such a dense stellar field.
Finally, in the vast majority of cases we do not detect any light from the lens
itself, the collected photons are from a background star completely unrelated
and distant from the lensing star. This is very different from other exoplanet
discovery methods and it means that we only obtain dynamical properties of
planets such their masses and periods.

@@

\section{Gravitational microlensing}
\label{sec:microlensing}

@@

>>>>

\section{Gravitational microlensing}
\emph{``Do not bodies act upon light at a distance and by their action bend its rays,
    and is not this action strongest at the least distance?''} asked \citet{Newton1704}.
Much later in 1911, even before he published his theory of General Relativity (GR), Einstein
considered the deflection of light from a distant star passing close
to a massive object in the foreground and derived an expression for the deflection angle of the light ray
which was off by a factor of 2 relative to the correct value.
He corrected the error in 1915 using GR.
The first experimental confirmation of this prediction came in 1919 during the
Solar eclipse when the astronomer Arthur Eddington famously measured the
deflection of light from distant stars passing close to the limb of the Sun
\footnote{Weather the data was actually good enough to make the claim that the
    deflection angle is equal to the one Einstein predicted is contested}.
Much later in 1936 a young amateur scientist R.W. Mandl convinced Einstein to
write a short a paper on the lensing effect by a massive star acting as the lens
instead of the Sun.
In the paper \citep{1936Sci....84..506E} Einstein derived an expression for the
magnification of the distant star when it is closely aligned to the foreground lens
star relative to an observer and he predicted that a luminous ring would form
if the two stars were perfectly aligned.
He considered this effect curious but useless, stating that
\emph{``Of course, there is no hope of observing this phenomenon directly.
    First, we shall scarcely ever approach closely enough to such a central line.
    Second, [the angles] will defy the resolving power of our instruments''}.

In this matter, Einstein was wrong. The famous astronomer Fritz Zwicky noticed
the phenomenon is likely to be observable at galactic scales if the lens was a
massive galaxy \citep{1937PhRv...51..290Z,1937PhRv...51..679Z} and that one
could use the measurement of the deflection angle to weigh the lens. Zwicky was
right and the first definitive observation came in 1979 by
\citet{1980ApJ...241..507Y} who observed a double image of the quasar Q0957+561
and concluded that the two images correspond to the same object whose light was
distorted by a massive galaxy. Many other observations followed, including
visible Einstein rings. Far from being just a curious phenomenon, galactic
scale lensing is now one of the key methods of observational cosmology, used
for inferring the parameters of the standard model and the distribution of dark
matter in the universe.

@@

That the same effect could be used to detect the presence of planets orbiting
around the lens star was first theorized by \citet{1964PhRv..133..835L} who
wrote that \emph{``the primary effect of planetary deflectors bound to stars
    other than the Sun will be to slightly perturb the lens action of these
    stars''} although he was also sceptical about the possibility of detection,
saying that \emph{``associated pulses would be so weak and infrequent and of
    such fleeting duration – perhaps a few hours – as to defy detection''}.
Gravitational lensing as a method for discovering exoplanets really took off
with the work of Paczyński
\citep{1986ApJ...304....1P,1986ApJ...301..503P,1991ApJ...374L..37M} who also
coined the term \emph{microlensing}, which refers to gravitational lensing in a
regime where the images of the background object cannot be resolved but one can
nevertheless measure its magnification as a function of time. Microlensing as a
method for detecting exoplanets has some unique aspects. First, it is a one-off
event which happens on a timescale of a few minutes up to several months
depending on the distances to the background star and the lensing star and the
mass of the lens. In addition to the fact that there's only one chance to
observe such an event for it to happen at all requires extremely precise
alignment between the lens and the background star and the chance of observing
a stellar microlensing event is about one-in-a-million (CITE) for a typical
star within the Milky Way. To observe a planetary signal is about an order of
magnitude less likely than that. Hence, obtaining a decent sample of planetary
events requires continuous monitoring on the order of $10^{8}$ stars at the
time which way microlensing observations that to focus on the densest region of
the Milky Way -- the galactic bulge. Figure~\ref{fig:crowded_field} a picture
from of such a dense stellar field. Finally, in the vast majority of cases we
do not detect any light from the lens itself, the collected photons are from a
background star completely unrelated and distant from the lensing star. This is
very different from other exoplanet discovery methods and it means that we only
obtain dynamical properties of planets such their masses and periods.

@@

\section{Gravitational microlensing}
\label{sec:microlensing}

@@

Gravitational lensing is a phenomenon fully described by Einstein's General
Theory of Relativity (GR) in which light changes its direction of propagation
when passing close to a massive body. GR predicts the following:
\begin{itemize}
    \item the presences of mass changes the spacetime geometry from the flat (Minkowski)
          metric $\eta_{\mu\nu}$ to a curved metric, specified by the tensor $g_{\mu\nu}$
    \item massless particles such as photons follow the null geodesics of the curved
          spacetime which are obtained by solving the \emph{geodesic equation}
\end{itemize}
If we restrict ourselves to the regime where the metric is time independent
and the particles are allowed to travel at any velocity less than $c$,
also known as the \emph{weak field} or \emph{linearized approximation }
of GR, the metric tensor
can be written as
\begin{equation}
    g_{\mu\nu}=\eta_{\mu\nu}+h_{\mu\nu}\quad,
\end{equation}
where $h_{\mu\nu}=
    \textrm{diag}(-2\Phi,-2\Phi,-2\Phi,-2\Phi)$\footnote{Assuming the
    the metric sign convention $(-,+,+,+)$} \citep{carroll_2019}.
The (static) Newtonian gravitational potential $\Phi$ obeys the Poisson equation
\begin{equation}
    \nabla^2\Phi=4\pi G\rho
\end{equation}
where $\rho$ is the mass density of the distribution of mass located between the observer
and the light source.
The geometry of the lensing system consisting of a distant source of light, the
observer, and a distribution of mass with density $\rho$ located between the
source and the observer is shown in Figure~\ref{fig:deflection_angle}.
\begin{figure}
    \centering
    \includegraphics[width=0.6\linewidth]{static/microlensing/deflection_angle.pdf}
    \caption{Geometry of gravitational lensing.
        A geodesic curve $x^\mu(\lambda)$ is deflected
        by an angle $\hat\alpha$ from its initial trajectory
        due to the presence of a massive body with mass density $\rho$. Figure
        adapted from \citet{carroll_2019}.}
    \label{fig:deflection_angle}
\end{figure}
A photon is deflected as it travels from a source to an observer by a
\emph{deflection angle} $\hat\alpha$, a vector in the
plane  perpendicular to the wave vector $\vec{k}$ pointing in the direction of
photon propagation.
The deflection angle is given by \citep{carroll_2019}
\begin{equation}
    \hat\alpha=2\int\nabla_\perp\Phi ds
    \label{eq:deflection_angle_general}
\end{equation}
where $\nabla_\perp\Phi$ is the gradient of the potential in the direction transverse
to the path of the photon and $s$ is the spatial distance traveled.
For a point mass $M$, the potential is given by the potential is
\begin{equation}
    \Phi=- \frac{GM}{(b^2+x^2)^{1/2}}
\end{equation}
where $x$ parametrizes the straight line distance between the observer and the lens and
$b$ is the impact parameter of the light ray.
The deflection angle (in SI units) is then given by
\begin{equation}
    \hat\alpha= \frac{4GM}{c^2b}= \frac{2R_s}{b}
\end{equation}
where the integral has been taken from $-\infty$ to $\infty$ (i.e. assuming that both
the source and the observer are located far away from the deflecting mass) and
$R_s=2GM/c^2$
is the Schwarzschild radius.
The deflection angle $\hat\alpha$ is directly proportional to the mass of the lens and
it does not depend on the wavelength of the light.
It also increases with the proximity of
the light ray to the lensing mass
which is the opposite behaviour to that of a
classical convex lens for which the deflection angle
vanishes at the center of the lens.
The same year Albert Einstein published his General theory of Relativity (1916),
English
physicist Sir Oliver Lodge suggested the light bending phenomenon could produce
a \emph{gravitational lens}.
He warned that it wouldn't be a lens in the usual sense,  because it does not have a
focal length.
In 1936, Einstein showed that a bright background light source perfectly aligned
with a foreground massive body would distort the source in the form of a ring
around the massive body.
However, he concluded that that there is no great chance of discovering such a
phenomenon.
Once astronomers started finding such ring like images in their data the name
\emph{gravitational lensing} caught on.

The resulting deflection angle is very small, for example, for the Sun we have
$GM/c^2=1.48\times 10^5\si{\centi\meter}$ (2.95 km), $R=6.96\times
    10^{10}\si{\centi\meter}$ and $\hat\alpha=1.75$ arc seconds. This deflection
effect was famously claimed to have been observed by Eddington when during a
total solar eclipse in 1919 \citet{2001ASPC..252...21C}, he observed the
position of background stars very close to the Sun's radius in the sky and
compared them to previous measurements taken when the stars weren't close to
the Sun in the sky. He concluded that the observed deflection was in agreement
with GR, although it is doubtful that his data is actually good enough to
distinguish between the GR prediction for the deflection of light and the
Newtonian prediction\footnote{ Gravitational bending on light can be predicted
    in Newtonian mechanics, assuming that light is a massless particle and using
    the equivalence principle which says that that all particles independent of
    their mass accelerate towards massive bodies.} , which is half of the GR value.

It is important to note that Eq.~\ref{eq:deflection_angle_general} is only
valid when the linearized approximation of GR is sufficient. Also, in this
approximation, the deflection angles are additive and the total deflection due
to several massive bodies is just the sum of the individual deflection angles.
The linearized approximation breaks down when the impact parameter $b$
approaches the Schwarzschild radius of the lensing mass and the linearized
metric is no longer sufficient. For most cases of astrophysical interest this
is not an issue.
